\documentclass[11pt,a4paper,twoside]{article}
\usepackage[utf8]{inputenc}
\usepackage{stylesheet}
\linespread{1.3}

\usepackage{import}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{bm}

\numberwithin{equation}{section}
\numberwithin{figure}{section}
\usepackage{natbib}
%\usepackage{biblatex}
%\usepackage[round]{natbib}  % formatting references as Name (Year) and (Name, Year)
% use parencite for all citations 
\renewcommand{\cite}[1]{\parencite{#1}}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{appendix}
%\usepackage[table,xcdraw]{xcolor}
\usepackage{blindtext}
\usepackage{url}

%try better bibliography formatting
\usepackage[style=authoryear, backend=biber]{biblatex}
\addbibresource{bibliography.bib}
\usepackage[nottoc]{tocbibind}
\setlength{\bibitemsep}{.5\baselineskip}

\captionsetup[table]{justification=centering,singlelinecheck=false} % table caption is left-aligned
%\bibliographystyle{plainnat} % reference list style

%\setlength\parindent{0pt} % remove indentation
%\setlength{\parskip}{1em} % add space between columns

%\import{}{Acronyms.tex}


\begin{document}

\maketitle\thispagestyle{empty}
\clearpage\pagenumbering{roman}

\section*{Abstract}
Mortality forecasts are important in e.g. demographical and actuarial sciences, and the Lee-Carter model \parencite{LeeCarter1992} is widely used for this purpose. Until now, it has not been possible to use the popular \inla methodology to perform Bayesian inference with this model. We show that the method proposed by \textcite{BachlLindgren2019}, implemented in the \texttt{R} library \inlabru, enables Bayesian inference with the \inla methodology on two versions of the Lee-Carter model for mortality projections. We first test this on synthetically produced data and then move on to test the procedure on real data for cancer mortality. In the first part of the study, we test the ability of the \inlabru method to correctly fit all parts of the Lee-Carter models to synthetic mortality rates produced by choosing arbitrary expressions for the different model components. We then assess the correctness of the resulting estimations. In the second part of the study, we use the \inlabru method to estimate current cancer mortality rates and predict future cancer mortality rates based on real data of cancer mortality, and assess the results. The results from both parts of the study indicate that the \inlabru method is a suitable tool for performing Bayesian inference on the Lee-Carter models. We compare the prediction results from the investigation on real data to results obtained with an equivalent estimation procedure using Age-Period-Cohort (APC) models. 

% Table of contents
\clearpage\tableofcontents
\newpage
\printacronyms

% Main body
\clearpage\pagenumbering{arabic}
\import{}{Report/1-Introduction.tex}
\import{}{Report/2-Theory.tex}
\import{}{Report/3-Synthetic-data.tex}
\import{}{Report/4-Real-data.tex}


% Reference list
\clearpage
\section*{References}
\addcontentsline{toc}{section}{References}
\printbibliography[heading=none]
%\bibliographystyle{apalike}
%\bibliographystyle{plainnat}
%\bibliography{Bibliography}

% Appendix
\begin{appendices}
\clearpage
\import{}{Report/Appendix.tex}
\end{appendices}

\end{document}